---
# MB5370: Introduction to Programming
title: "Workshop 1 Module 1"
author: "Ryan Waln"
date: "2026-05-20"
output: html_document
Workspace: D:/Tri 2 2026/Techniques in Marine Science 1/Programming & Data Analysis/Code/Basic_Code_M1W1.Rmd
---

# download and install then comment out so it doesn't run again
```{r}
install.packages("tidyverse") 
#install.packages("redlistr") 
#install.packages("terra") 
```

# load packages
```{r}
library("ggplot2")
library("tidyr")
# top 2 are included in tidyverse library
library("tidyverse") 
```

#Help options
```{r}
?round # go to help
args(round) # use args in the Console
```

#Basics
```{r}
2+1
1:30
6 * 2
6 % 2
Age <- 25 
first_name <- 'Bill'
Age + 1
Age + Age 
x = 15 + 25.1 + 20.25
y = 5
x + y
```

# Naming Errors
```{r}
01_age <- 25 # starts with a number
!_age <- 25 # no special symbols
age bob <- 25 # no spaces
`age bob` <- 25 # no spaces, but with back ticks
```

#Functions 
```{r}
years_old <- 25.7
round(years_old) # rounds up
floor(years_old) # rounds down
years_old <- 25.765
round (years_old, 2) # comma after the object to specify argument to 2 decimal places
```

#Paste Function
```{r}
years_old <- 24
Ryan <- paste("Ryan is", years_old)
```

# Misconceptions (variables in programs do not work the same way as they do in spreadsheets)
```{r}
grade <- 55
total <- grade + 10
print (total)
grade <- 90
print (total) # value of total in a spreadsheet will be 100, but in programming a variable holds the value it was assigned (65)
total <- grade + 10
print (total) # executed in the way it was defined

```

# Debugging errors
```{r}
p <- 2
z <- 5
out <- p * z  # What should the value of out be?
print (out) # What is the value of out? Is it the same as expected?
out <- p * a # undefined variable.
## > Error: object 'a' not found
out <- p * z
```

```{r}
#Lacks commas between objects
my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro"
             "cinco")
print (my_quis) #Quiz misspelled
str(my_quiz)
len(my_quiz) #Len should be length

#Fixed code
my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro",
             "cinco")
print (my_quiz)
str(my_quiz)
length(my_quiz)
```

#Testing if data is numeric or a character
```{r}
x <- 1
is.character(x)
is.numeric (x)
```

#Types of Data
```{r}
#Vector
y <- c(1, 2, 3)
z <- c("Sarah", "Tracy", "Jon")

class(y) 
class(z)

#Lists (values inside can be of different types)
x <- list(1, "a", TRUE)
x
x[[2]] #Retrieve 2nd value

#Insted of importing .csv files you can make your own data frames by adding vectors and giving them column names
my_dataframe <-  data.frame (no = c(1,2,3), name = c("Tracey", "John", "Pete"), pass = c(TRUE, FALSE, TRUE))
my_dataframe
str(my_dataframe)

#Set 1st column as categorical rather than numeric 
my_dataframe$no = as.factor(my_dataframe$no)
str (my_dataframe)
```

#------------------------------------#
#Class discussion. We want to take a .csv dataset representing counts from a BRUVs survey and turn it 
#into a plot of the abundance of a single species. What major steps - the main tasks you would need to do using programming 
#might be required to get from a structured .csv table to a plot of a single species? Record your #answers in your script using numbered comments. 


# 1. Read the .csv dataset into R. etc.
# 2. Sort the data as needed and define NA
# 3. Plot the data selected with the plot function and edit abundance plot as desired
#------------------------------------#
