---
# MB5370: Introduction to Programming
title: "Workshop 4 Module 1"
author: "Ryan Waln"
date: "2026-05-27"
output: html_document
---

## Introduction

This file covers:
- How to set up AI assistants like GitCopilot and chattr AI to assist with coding
- Sorting data with conditional statements 
- How to set up classic loops
- How to set up Purrr loops
- How to use functions

```{r housekeeping, message=FALSE, warning=FALSE, include=FALSE}
# R Setup & housekeeping
knitr::opts_chunk$set(
	echo = TRUE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)

rm(list=ls())
objects()
```

## Setting Up AI assistants

#### Set up GitCopilot in Tools > Global Options > Assistant

#### Install chattr AI and other packages
```{r}
#install.packages("ellmer")
#install.packages("chattr")
#install.packages("palmerpenguins")
```

```{r library}
library("knitr")
library(usethis)
```

#### Test git copilot
```{r}
# Load tidyverse and palmerpenguins libraries 
library("tidyverse")
library("palmerpenguins")
```

```{r}
# Create a ggplot scatter plot using the penguins dataset. 
# Plot bill_length_mm on the x-axis and bill_depth_mm on the y-axis.
# Color points by species and use geom_point with size 3.
# Add a minimal theme and clean labels.   
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Penguin Bill Dimensions",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species")

```
```{r}
# Create boxplot using penguins dataset 
# Plot flipper_length_mm across different islands
# Color by species and use geom_point with size 3.
# Add a minimal theme and clean labels.  
ggplot(penguins, aes(x = island, y = flipper_length_mm, color = species)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.2), size = 3) +
  theme_minimal() +
  labs(title = "Penguin Flipper Length by Island",
       x = "Island",
       y = "Flipper Length (mm)",
       color = "Species")
```

#### Link chattr AI to Google Gemini LLM
```{r}
## run in console
# usethis::edit_r_environ()
## Enter into new window
# GEMINI_API_KEY=your_actual_key_here
## Restart R session for changes to take effect

## Ensure R uses prefered model each session
usethis::edit_r_profile()
## Enter into new window
options(.chattr_chat = ellmer::chat_google_gemini())
## Restart R session for changes to take effect
```

#### Check API Key is Recognized
```{r}
Sys.getenv("GEMINI_API_KEY")
```

## Set up chat
```{r}
library("chattr")
library("ellmer")
library(ggplot2)
my_chat <- ellmer::chat_google_gemini(model = "gemini-2.5-flash")
chattr_use(my_chat)
chattr::chattr("say hello")
  
```

## Start a Chat
```{r}
chattr_app()
#Execute Key shortcut: Ctrl + Alt + c
```

#### Test chattr AI with ggplot2 prompt: "Show me how to load the built-in iris dataset and use ggplot2 to build a scatter plot of Sepal.Length versus Sepal.Width colored by Species."
```{r}
data(iris)

ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
  geom_point()
```


## Sorting Data


### ifthen statments:if a condition is meet, do this command.

#### Change object if it equates to something
```{r}
x <- 1
x # what is x?
if (x == 1) {
  x <- 2
}
x # what is x now?

x <- 3
x # what is x?
if (x == 1) {
  x <- 2
}
x # what is x now?
```

### ifelse staments: used to sort data, such as if a conditon is not met start a second set of commands.

#### Example:
if (animal_data_received) {
	1. save statement in log file that the animal has been updated
} else {
	1. save statement saying no data received
}


### elseif statements: allows additional conditional commands between the if and else

#### Example of ifelse code:
```{r}
name <- "Nick"
if (name == "Nick"){
  last.name <- "Murray"
}
print (paste(name, last.name)) 
# try changing name to Bob, what happens?

# now handle that case
name <- "Bob"
if (name == "Nick"){
  last.name <- "Murray"
} else {
  last.name <- "Jones"
}
print (paste(name, last.name))

# chaining together
name <- "Jim"
if (name == "Nick"){
  last.name <- "Murray"
} else if (name == "Bob") {
  last.name <- "Jones"
} else {
  last.name <- "... last name is unknown"
}
print (paste(name, last.name))

```

#### Simple Example: Checking Sea Surface Temperature (SST)
```{r}
#() set the logical condition and curly braces {} used for the action if that condition is true
sst <- 30.2

if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
} else {
  print("SST remains within baseline parameters.")
}

```

### if_else()
#### The if_else(condition, true_output, false_output) function evaluates an entire column item-by-item:
```{r}
# Sample coral data
coral_monitoring <- tibble(
  site = c("Site_A", "Site_B", "Site_C"),
  depth_m = c(12, 35, 8)
)

# Classify depth using if_else
coral_monitoring <- coral_monitoring %>% 
  mutate(zone = if_else(depth_m > 30, "Deep Reef", "Shallow Reef"))

coral_monitoring


## %>%: used to chain functions together (passes the result of the expression on its left as the first argument to the function on its right)
## Mutate: used to edit columns in a data frame
```


### case_when()
#### When you have more than two possibilities, stacking multiple if-else blocks becomes unreadable. case_when() is better and uses formula tildes (~):
```{r}
coral_monitoring <- coral_monitoring %>% 
  mutate(reef_category = case_when(
    depth_m < 10  ~ "Lagoon / Flats",
    depth_m <= 30 ~ "Crest / Slope",
    depth_m > 30  ~ "Mesophotic / Deep",
    TRUE          ~ "Unclassified" # Catch-all remainder
  ))

```

```{r}
# create a tibble named marine_stations containing a column for salinity with values 35, 28, 32, and 12
# Use the pipe operator (%>%) and mutate() combined with case_when() to create a new column named environment_type
# Classify values below 15 as "Estuarine", values between 15 and 30 as "Brackish", and values above 30 as "Marine"
marine_stations <- tibble(
  salinity = c(35, 28, 32, 12)
)
marine_stations <- marine_stations %>% 
  mutate(environment_type = case_when(
    salinity < 15 ~ "Estuarine",
    salinity >= 15 & salinity <= 30 ~ "Brackish",
    salinity > 30 ~ "Marine"
  ))

marine_stations
```


## Classic Looping


#### Create a new object for every loop run with a name value
```{r}
for (value in c("My", "first", "for", "loop")) {
  print("one run")
}

```

#### Print out values for each loop instead of a given label
```{r}
for (value in c("My", "second", "for", "loop")) {
  print(value)
}
```

#### Can use any word you want with print() 
```{r}
for (word in c("My", "second", "for", "loop")) {
  print(word)
}
for (string in c("My", "second", "for", "loop")) {
  print(string)
}
for (i in c("My", "second", "for", "loop")) {
  print(i)
}


```

#### Classic loop anatomy
```{r}
for (year in 2020:2024) {
  print(paste("Processing climate data for year:", year))
}
```

#### Loop across index sequences
```{r}

transect_lengths <- c(50, 100, 25, 75)

for (i in seq_along(transect_lengths)) {
  print(paste("Transect number", i, "measures", transect_lengths[i], "metres."))
}
```

#### Make your own for loop: A list of your top 3 favourite animals
```{r}
#Make a for loop that lists my top 3 favorite animals using the print() function and paste() to combine text and the loop value

favortie_animals <- c("Snow Leopard", "Octopus", "Sperm Whale")
My_number <- c(1, 2, 3)
for (i in seq_along(My_number)) {
  print(paste ("My number",i, "favorite animal is the", favortie_animals[i] ))

  }

```

#### Create list to be looped and compactly display it
```{r}
my_list <- 1:50 
print (my_list)
str(my_list)
```

#### Run the loop
```{r}
# make a for loop (i is typically used)
for(i in my_list) {                              # Head of for-loop
  result <- i
  out <- paste0("I am printing loop number: ", i, ".")# Output: 1 loop
  print(out)                                    # Using print function
}  # end of loop
print ("For loop complete!")
print ("Program complete!")

```

#### Add a number to every item in the list being looped
```{r}
add_me <- 500     #Establish added value
for(i in my_list) {                              # Head of for-loop
  result <- i + add_me # add to the i value
  out <- paste0("Result of iteration: ", i, " = ", result, ".") 
  print(out)    
}   #End of loop
print ("For loop complete!")
print ("Program complete!")

```

#### Code for when you need to extract a value from a data file and add it to another one and so on
```{r}
# Note we have i and j for each loop
for (i in c(1, 2, 3, 4, 5)) {
  print (paste ("I am base loop number: ",i))
  Sys.sleep(0.8) # slow it down for example purposes
  for (j in c(1, 2, 3)){ 
    print (paste ("... I am inside loop number: ",j))
  }
}

```

#### While loops: continue to loop while a condition remains true
#### Repeat loops: continue until you tell them to stop


## Purrr Looping


### Purrr Functions 
#### map(): applies a function to each element of a list or vector and returns a list of the same length
#### map_dbl(): applies a function to each element of a list or vector and returns a double (numeric) vector
#### map_chr(): applies a function to each element of a list or vector and returns a character
#### map_df(): applies a function to each element of a list or vector and returns a data frame


#### Standard for Loop to change values on list with sqrt function vs...
```{r}
site_areas <- c(144, 400, 625)
results <- numeric(length(site_areas)) # Must build an empty container first

for(i in seq_along(site_areas)) {
  results[i] <- sqrt(site_areas[i])
}
results
```

#### ...purrr equivalent
```{r}
library(purrr)
# Map directly and define your expected output type explicitly
mapped_results <- map_dbl(site_areas, sqrt)
mapped_results
```

#### Combine purrr pipelines inside data frames to split analysis across separate groups
```{r}
# typical summary
summary(iris) 

# Use pipeline to split data into species
iris %>%  
  split(.$Species) %>%  # Splits the iris data set into a list of data frames based on the Species column (insert iris into dot)
  map(~summary(.x)) # Finds summary of species

```

#### Using Chattr to compare methods when calculating means for fish count data
```{r}
fish_counts <- list(
  river_a = c(12, 15, 10, 18, 13),
  river_b = c(25, 22, 28, 30, 24),
  river_c = c(8, 9, 7, 10, 6)
)

# Using a for-loop
mean_counts_for_loop <- numeric(length(fish_counts))
for (i in seq_along(fish_counts)) {
  mean_counts_for_loop[i] <- mean(fish_counts[[i]])
}
print(mean_counts_for_loop)

# Using purrr::map_dbl
mean_counts_purrr <- map_dbl(fish_counts, mean)
print(mean_counts_purrr)
```


## Functions

#### Functions are useful for avoiding repeated code

#### Function consists of a Name, Arguments (the input variables), and a Body (the execution logic wrapped inside curly braces)

####  Good practice to only use arguments passed directly into them, rather than accidentally grabbing random background objects in your global R environment
```{r}
# Define Function
calculate_coral_mortality <- function(initial_count, surviving_count) {
  
  # Logic safety switch using our conditional tools!
  if (initial_count <= 0) {
    stop("Initial count must be greater than zero.")
  }
  # Calculate mortality rate
  mortality_rate <- (initial_count - surviving_count) / initial_count
  return(mortality_rate)
}

# Test custom function
calculate_coral_mortality(initial_count = 120, surviving_count = 84)
```
#### Other Examples include converting temperature units: temp_C <- (temp_F - 32) * 5 / 9

#### Can use a function to do this for every value in a dataset
```{r}
fahrenheit_to_celsius <- function(temp_F) {    # establish conversion as output of function operating on the arguments
   # convert F to C (good practice to always list what function does)
   temp_C <- (temp_F - 32) * 5 / 9
  return(temp_C)                     
}

# run the function on 2 lines
fahrenheit_to_celsius(100)
fahrenheit_to_celsius(200)
```

#### Celsius to fahrenheit
```{r}
celsius_to_fahrenheit <- function(temp_C) {    
   # convert C to F
   temp_F <- (temp_C * 9 / 5) + 32 
  return(temp_F)                     
}

celsius_to_fahrenheit(0)
celsius_to_fahrenheit(20)
celsius_to_fahrenheit(40)
```

#### Temp conversion w/ saftey stop for absolute zero
```{r}
convert_temp_c_to_f <- function(temp_C) {
  if (temp_C < -273.15) {
    stop("Temperature cannot be below absolute zero (-273.15°C).")
  }
  temp_F <- (temp_C * 9 / 5) + 32
  return(temp_F)
}

convert_temp_c_to_f(25)
#convert_temp_c_to_f(-300) # Triggers the safety check and stops execution with an error message
```

#### Advanced problems
```{r}
# Create custom function named clean_and_classify_survey that accepts a data frame and a column name threshold parameter as inputs
# Safely check if the specified column name exists within the passed data frame; if it does not, trigger a user-friendly error message using stop()
# Filter out rows containing missing values (NA) within that specific column
# Use the pipe operator and mutate() along with a vectorised conditional (if_else() or case_when()) to append a status string variable classification based on your threshold parameter.
# Return the cleaned, altered data frame to the global environment.
# Test your custom pipeline using a mock tibble that contains at least one missing value to confirm the code handles anomalies without crashing.

clean_and_classify_survey <- function(df, column_name, threshold) {
  # Check if the specified column exists in the data frame
  if (!column_name %in% colnames(df)) {
    stop(paste("Error: Column", column_name, "not found in the data frame."))
  }
  
  # Filter out rows with NA values in the specified column and classify based on threshold
  cleaned_df <- df %>%
    filter(!is.na(.data[[column_name]])) %>%
    mutate(status = if_else(.data[[column_name]] > threshold, "Above Threshold", "Below Threshold"))
  
  return(cleaned_df)
}
# Test the function with a mock tibble
test_data <- tibble(
  id = 1:10,
  measurement = c(5, 7, NA, 3, 8,
                 2, 9, NA, 4, 6)
)
result <- clean_and_classify_survey(test_data, "measurement", threshold = 5)
print(result)
```

```{r}
#Given a vector of raw survey counts counts <- c(15, 24, 8, 42) and a vector of species-specific scaling factors factors <- c(1.2, 0.8, 2.5, 1.1), write a mapping statement that iterates across both vectors simultaneously to return a single vector of scaled abundance values. Do not use a manual index loop.

library(purrr)

counts <- c(15, 24, 8, 42)
factors <- c(1.2, 0.8, 2.5, 1.1)

scaled_abundance <- map2_dbl(counts, factors, ~ .x * .y)
print(scaled_abundance)

```

