---
# MB5370: Introduction to Programming
title: "Data_ Looping_Module01_Workshop03"
author: "Ryan Waln"
date: "2026-05-21"
output: html_document
Workspace: E:/Tri 2 2026/Techniques in Marine Science 1/Programming & Data Analysis/Code/GitHub_Looping_Data_M1W3.Rmd
---
```{r install packages}
#install.packages("tidyverse") 
#install.packages("knitr")
#install.packages("usethis")
```

```{r remove packages}
#remove.packages()
```

```{r library}
library("tidyverse") 
library("knitr")
library(usethis)
```

# Github setup (run in console)
```{r}
##Get git credentials
#credentials::git_credential_ask()  

## see Git status
#usethis::git_sitrep()

## create Personal Access Token
#usethis::create_github_token() 

## prompt to insert PAT
#gitcreds::gitcreds_set() 

##track current project locally (can also use Tools > Project Options … > Git/SVN, and set to Git)
## or run  git init in terminal
#usethis::use_git() 

##Enter into terminal if using USB
#git config --global --add safe.directory "E:/Tri 2 2026/Techniques in Marine Science 1/Programming & Data Analysis/Code"  

##Then restart R and run 
#usethis::git_sitrep() 
```

# Workflow for github:
## Edit → Save → Stage → Commit → Pull → Push
```{r}
#usethis::use_github() # Have github start tracking files
```


---------------------------------------------------------------------------------------------------------------------------------------------------------------------
# Looping
---------------------------------------------------------------------------------------------------------------------------------------------------------------------
## Create a new object for every loop run with a name value
```{r}
for (value in c("My", "first", "for", "loop")) {
  print("one run")
}

```

## Print out values for each loop instead of a given label
```{r}
for (value in c("My", "second", "for", "loop")) {
  print(value)
}
```
## Can use any word you want with print() 
```{r}
for (word in c("My", "second", "for", "loop")) {
  print(word)
}
for (string in c("My", "second", "for", "loop")) {
  print(string)
}
for (i in c("My", "second", "for", "loop")) {
  print(i)
}


```

## Create list to be looped and compactly display it
```{r}
my_list <- 1:50 
print (my_list)
str(my_list)
```

## Run the loop
```{r}
# make a for loop (i is typically used)
for(i in my_list) {                              # Head of for-loop
  result <- i
  out <- paste0("I am printing loop number: ", i, ".")# Output: 1 loop
  print(out)                                    # Using print function
}  # end of loop
print ("For loop complete!")
print ("Program complete!")

```

## Add a number to every item in the list being looped
```{r}
add_me <- 500     #Establish added value
for(i in my_list) {                              # Head of for-loop
  result <- i + add_me # add to the i value
  out <- paste0("Result of iteration: ", i, " = ", result, ".") 
  print(out)    
}   #End of loop
print ("For loop complete!")
print ("Program complete!")

```

## Code for when you need to extract a value from a data file and add it to another one and so on
```{r}
# Note we have i and j for each loop
for (i in c(1, 2, 3, 4, 5)) {
  print (paste ("I am base loop number: ",i))
  Sys.sleep(0.8) # slow it down for example purposes
  for (j in c(1, 2, 3)){ 
    print (paste ("... I am inside loop number: ",j))
  }
}

```

## While loops: continue to loop while a condition remains true
## Repeat loops: continue until you tell them to stop


## PINNED Make your own for loop. Make a list of your top 3 favourite animals
```{r}
for (i in c(1, 2, 3)) {
  print (paste ("My number",i))
  for (j in c("Snow Leopard", "Octopus", "Sperm Whale")){ 
    print (paste ("...favorite animal is the",j))
  }
}
```
---------------------------------------------------------------------------------------------------------------------------------------------------------------------
# Ifthen & Elseif
---------------------------------------------------------------------------------------------------------------------------------------------------------------------

## ifthen statments:if a condition is meet, do this command.

## Change object if it equates to something
```{r}
x <- 1
x # what is x?
if (x == 1) {
  x <- 2
}
x # what is x now?

x <- 3
x # what is x?
if (x == 1) {
  x <- 2
}
x # what is x now?
```

## ifelse staments: used to sort data, such as if a conditon is not met start a second set of commands.

if (animal_data_received) {
	# 1. save statement in log file that the animal has been updated
} else {
	# 1. save statement saying no data received
}

## elseif statements: allows additional conditional commands between the if and else

## Example of ifelse code
```{r}
name <- "Nick"
if (name == "Nick"){
  last.name <- "Murray"
}
print (paste(name, last.name)) 
# try changing name to Bob, what happens?

# now handle that case
name <- "Bob"
if (name == "Nick"){
  last.name <- "Murray"
} else {
  last.name <- "Jones"
}
print (paste(name, last.name))

# chaining together
name <- "Jim"
if (name == "Nick"){
  last.name <- "Murray"
} else if (name == "Bob") {
  last.name <- "Jones"
} else {
  last.name <- "... last name is unknown"
}
print (paste(name, last.name))

```
---------------------------------------------------------------------------------------------------------------------------------------------------------------------
# Functions
---------------------------------------------------------------------------------------------------------------------------------------------------------------------
## Functions are useful for avoiding repeated code

## Examples include converting temperature units: temp_C <- (temp_F - 32) * 5 / 9

## Can use a function to do this for every value in a dataset
```{r}
fahrenheit_to_celsius <- function(temp_F) {    # establish conversion as output of function operating on the arguments
   # convert F to C (good practice to always list what function does)
   temp_C <- (temp_F - 32) * 5 / 9
  return(temp_C)                     
}

# run the function on 2 lines
fahrenheit_to_celsius(100)
fahrenheit_to_celsius(200)
```
## Celsius to fahrenheit
```{r}
celsius_to_fahrenheit <- function(temp_C) {    
   # convert C to F
   temp_F <- (temp_C * 9 / 5) + 32 
  return(temp_F)                     
}


celsius_to_fahrenheit(0)
celsius_to_fahrenheit(20)
celsius_to_fahrenheit(40)
```

-----------------------------------------------------------------------------------------------------------------------------------------------------------------

#Execute other r script such as workshop 1 R script (doesn't work with R markdown files)
```{r}
source("../Code/Workshop_1.R")
```


