---
# MB5370: Introduction to Programming
title: "Workshop 2 Module 1"
author: "Ryan Waln"
date: "2026-05-21"
output: html_document
---

## Introduction

This file covers basic plotting functions with ggplot2. Material covered includes:
- Plotting by category by color, shape, etc. 
- Plotting multiple geoms 
- Plotting and highlighting subsets of data
- Transforming data with stat_count() to determine best fit plot type 
- Aesthetic adjustments 
- Plot template at end

```{r housekeeping, message=FALSE, warning=FALSE, include=FALSE}
# R Setup & housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)

rm(list=ls())
objects()
```

#### Shortcuts
Ctrl + Alt + I #creates new chunks
Ctrl + Enter #to run selected line
Ctrl + Shift + Enter #to run whole chunk
 
when kniting: more # = smaller text on printed documents

```{r install packages}
# install.packages("tidyverse") 
# install.packages("knitr")
```

```{r remove packages}
# remove.packages()
```

```{r library}
library("tidyverse") 
library("knitr") 
```


## Plotting


#### Create plot with mpg dataset from tidyverse (can use cursor and tab to select x & y varriables)
```{r}
mpg
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) 
#displ is engine size, hwy is fuel efficency
#geom_points creates scatter plot, aes() is used to select x and y variables
```

#### Plot template using ggplot()
```{r ggplot template}
#ggplot(data = <DATA>) + 
#  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>)
```

#### Set car class to be displayed by color using aes()
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))

```

#### Set car class to be displayed by size of point (not good data visualization)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))

```

#### Set car class to be displayed by different transparency
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

```

#### Set car class to be displayed by different shapes (not recomended when > 6 categories)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))

```

#### Turn all points blue 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")

```

#### Color code engine sizes > than 5 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = displ < 5))
```

#### Error reminder: the + must be on top line
```{r}
ggplot(data = mpg) 
+ geom_point(mapping = aes(x = displ, y = hwy)) 


```

#### Creating sub plots/pannels. Divide by class and set 2 rows with facet_wrap()
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2) #~ for object, nrow for no. of rows
```

#### Use facet_grid() to split up plots by multiple variables (separate variables w/ ~)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

#### Use . if you don't want to include the rows or columns section
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))+ 
facet_grid(. ~ cyl)

```

#### Plot data as smooth line w/ geom_smooth()
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

#### Can comment out plots that don't display data well to keep track
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy)) # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # smooth line better

```

#### Can generate mutiple lines for different categories with linetype = <insert>
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))

```

#### Use group = <insert> to sort data by a category
```{r}
 ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

#### Change color of each line based on drv and exclude legend
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )


```

#### Plot data as sacterplot and smooth line graph on top of each other (would have to change x variable in both locations to edit)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

#### More efficent way to plot above set if you want to change x & y varriables
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

#### Plots are the same just different methods
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))

```

#### Can edit one geom to separate things by color
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()

```

#### Plot a subset of the data with one geom (only subcompact vehicles)
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

#### Change geom_(line, histogram, box, bar) to get the desired chart type
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_line()
```

## Transforming Data (diamonds dataset)

#### Bar chart shows most have high quality cuts
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

#### stat_count() can be used instead of geom (automatically does stats and best fit plot)
```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))

```

#### Change count (a summary) to identify which uses raw values. (Required for when you want to change defaults ) 
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")


```

## Aethstetic Adjustments

#### fill =  can add differential colors to bar chart on diamond cut or used to map proportion of cuts with certian clarity
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

```

#### Position Addjustments: identity (raw data), fill (changes heights), and dodge (forces ggplot2 to not put things on top of each other)

#### To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")   #Identity uses raw data points instead

ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 3/5, position = "identity")
```

#### To color the bar outlines with no fill color (never use)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")

```

#### To make bars the same height to look at proportions in a group use: position = "fill"
#### Can do something similar with position = "doge" to place objects besides each other (pop-size structure)
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

#### position = "jitter" adds a small amount of random noise to each point lessen points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```


# PLOT TEMPLATE (fill in <>)

```{r}
#ggplot(data = <DATA>) + 
#  <GEOM_FUNCTION>(
#     mapping = aes(<MAPPINGS>),
#    stat = <STAT>, 
#     position = <POSITION>
#  ) +
# <FACET_FUNCTION>

```






